---
title: "Lab 11"
author: "Summer Heschong"
date: "2025-04-02"
output: pdf_document
---
# Setup
```{r setup}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(parameters)

#load data
cocos <- read.csv(here('Data/Raw/palmyra_cocos.csv'))
```

# Factors
```{r factors}
#examine data structure
str(cocos)

#make all text columns into factors while specifying group order
cocos <- cocos %>%
  mutate(Type = factor(Type, levels = c('Mono', 'Mixed')),
         Rats = factor(Rats, levels = c('pre', 'post')),
         Plot = factor(Plot, levels = c('Eastern', 'Holei 1', 'Kaula',
                                        'Papala', 'Paradise', 'Sand',
                                        'Holei 2')))

```

# Multi-level Regresssion
## Random Intercepts
### Step 1- Define Research Question
Do annual precipitation, forest type, and eradication treatment significantly predict percent cover of coconut tree seedlings? 

### Step 2 - Examine data and possible correlations
```{r display}
#show distributions of raw data
ggplot(cocos, aes(x = Ppt)) +
  geom_histogram(color = 'black')

ggplot(cocos, aes(x = Type)) +
  geom_bar()

ggplot(cocos, aes(x = Rats)) +
  geom_bar()

ggplot(cocos, aes(x = perc_CN)) +
  geom_histogram(color = 'black')

#show distributions of predictor variables over random effect variable
ggplot(cocos, aes(x = Plot, y = Ppt)) +
  geom_boxplot()

ggplot(cocos, aes(x = Plot, y = Type)) +
  geom_col(position = 'dodge') 

ggplot(cocos, aes(x = Plot, y = Rats)) +
  geom_col(position = 'dodge', color = 'black') 

#show relationships between values
ggplot(cocos, aes(x = Ppt, y = perc_CN)) +
  geom_point()

ggplot(cocos, aes(x = Type, y = perc_CN)) +
  geom_point()

ggplot(cocos, aes(x = Rats, y = perc_CN)) +
  geom_point()
```

### Step 3 - Fit linear regresion model and evaluate model diagnostics
```{r lrm}
#fit linear regression model
cocos.lm1 <- lm(perc_CN ~ Ppt + Type + Rats, data = cocos)

#examine results
summary(cocos.lm1)
plot(cocos.lm1)
```

### Step 4 - Fit linear regression model with random effect and evaluate diagnostics
```{r multilevel}
# refit standard linear regression with gls
cocos.lm2 <- gls(perc_CN ~ Ppt + Type + Rats, data = cocos)

#fit a multi-level model
cocos.lmm1 <- lme(perc_CN ~ Ppt + Type + Rats, 
                  random = ~ 1|Plot, data = cocos)

#compare the two models
AIC(cocos.lm2, cocos.lmm1)

#examine model rsutls
plot(cocos.lmm1)
qqnorm(cocos.lmm1)
summary(cocos.lmm1)

```

### Step 5 - Communicate methods and results
a. describe model
b. plot results
```{r results}
#sample report visualizatioin
fig1 <- ggplot(cocos, aes(x = Rats, y = perc_CN)) +
  geom_boxplot(width = 0.2) +
  geom_jitter(aes(color = Plot), size = 3, width = 0.10) +
  labs(x = 'Rat Eradication Status',
       y = 'Percent C. nucifera Seedlings') +
  theme_bw()

fig1

#output results as table
tbl_regression(cocos.lmm1)

```

## Nested Random Intercepts
### Skipping to Step 4 - Fit multilevel model
```{r nested}
# first add location column
cocos <- cocos %>%
  mutate(Side = factor(case_when(Plot %in% c("Eastern",
                                             'Holei 1',
                                             'Holei 2') ~ 'East',
                                                   TRUE ~ 'West'),
                       levels = c('East', 'West'))) 

#fit model (when nesting the larger grouping should be above the smaller)
cocos.lmm2 <- lme(perc_CN ~ Ppt + Type + Rats, 
                  random = ~ 1 | Side / Plot, data = cocos)

#compare to standard regression
AIC(cocos.lm2, cocos.lmm2)

#Examine results
plot(cocos.lmm2)
qqnorm(cocos.lmm2)
summary(cocos.lmm2)
```

## Random Slopes
### Start from Step 4 - fit multi level model 
```{r slopes}
# fit a multi level model 
cocos.lmm3 <- lme(perc_CN ~ Ppt + Type + Rats, 
                  random = ~ 1 + Ppt| Plot, data = cocos)

#compare to standard model
AIC(cocos.lm2, cocos.lmm3)

# Examine results
plot(cocos.lmm3)
qqnorm(cocos.lmm3)
summary(cocos.lmm3)

#examine what slopes and intercepts look liker per plot
coef(cocos.lmm3)

#save each plots equation including the average line
eqAvg <- function(x){0.1170558 + -0.19002*x}
eqEastern <- function (x){ 0.8311340 + -8.654819e-06 *x}
eqHolei1 <- function(x){0.6414776 + 2.907469e-06*x}
eqKaula <- function(x){0.5897089 + -8.956022e-06*x}
eqPapala <- function(x){0.9229401+ -2.291900e-05*x}
eqParadise <- function(x){0.9562606 +-3.197096e-05*x}
eqSand <- function(x){0.7360257 +-2.158791e-05*x}
eqHolei2 <- function(x){0.5740899 +-5.450407e-05*x}

#plot them together
fig2 <- ggplot(cocos, aes(x = Ppt,
                          y = perc_CN)) +
  geom_point(aes(color = Plot), size = 3)+ 
  labs(x = 'Cumulative Annual Precip(mm)',
       y = 'Percent C. nucifera Seedlings') +
  stat_function(fun = eqEastern, color = 'black') +
   stat_function(fun = eqHolei1, color = 'blue') +
   stat_function(fun = eqKaula, color = 'red') +
   stat_function(fun = eqPapala, color = 'yellow') +
   stat_function(fun = eqParadise, color = 'orange') +
   stat_function(fun = eqSand, color = 'green') +
   stat_function(fun = eqHolei2, color = 'purple') +
   stat_function(fun = eqAvg, color = 'maroon') +
  theme_bw()
fig2
```


