---
title: "Lab9"
author: "Summer Heschong"
date: "2025-03-19"
output: pdf_document
---
# setup
```{r setup}
library(here)
library(tidyverse)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)

#load data
sharkriver <- read_csv(here('Data/Raw/fce_sharkriver.csv'))
coralfish <- read_csv(here('Data/Raw/mcr_fish.csv'))

```

# Logistic Regression

```{r, tidy}
sharkriver2 <- sharkriver %>%
  replace_with_na_all(condition = ~.x == -9999) %>%
  filter(DO <30)

```

## Step 1 define research question
Do water temperature, salinity, and season significantly predict hypoxic conditions?

```{r tidy more}

#make columns for hypoxia and seasons
sharkriver2 <- sharkriver2 %>%
  mutate(hypoxia = case_when(DO <= 4 ~ 'YES',
                             DO > 4 ~ 'NO')) %>%
  mutate(season = case_when(DayofYear >= 59 & DayofYear < 151 ~ 'Spring',
                            DayofYear >= 151 & DayofYear < 243 ~ 'Summer',
                            DayofYear >= 243 & DayofYear < 334 ~ 'Fall',
                            TRUE ~ 'Winter'))


```

## Step 2 examine data
### a. display distributions or raw counts of data
```{r vis data}
#create histograms of data

ggplot(sharkriver2, aes(x = hypoxia)) +
geom_bar()

ggplot(sharkriver2, aes(x = WaterTemp)) +
geom_histogram(color = 'black')

ggplot(sharkriver2, aes(x = Salinity)) +
geom_histogram(color = 'black')

ggplot(sharkriver2, aes(x = season)) +
geom_bar(color = 'black')



```

### b. display relationshipbs between predictor and outcome

```{r vis data more}
#create scatterplots

ggplot(sharkriver2, aes(x = WaterTemp, y = hypoxia)) +
geom_point(color = 'black')

ggplot(sharkriver2, aes(x = Salinity, y = hypoxia)) +
geom_point(color = 'black')

ggplot(sharkriver2, aes(x = season, y = hypoxia)) +
geom_point(color = 'black')



```

## Step 3 fit regression model

```{r factors}
#change chr to factor
sharkriver2$hypoxia <- factor(sharkriver2$hypoxia,
                              levels = c('NO', 'YES'))

sharkriver2$season <- factor(sharkriver2$season,
                             levels = c('Winter', 'Spring',
                                        'Summer', 'Fall'))
#fit logistic regressioin model
hypox_mod <- glm(hypoxia ~ WaterTemp + Salinity + season,
                 data = sharkriver2,
                 family = 'binomial')

```

## Step 4 evaluate model diagnostics

```{r review}
#examine model output
summary(hypox_mod)

#examine modified residuals plot
simulateResiduals(hypox_mod) %>% plot()

```

#Step 5 - Interpret model and communicate the results
```{r vis extra}
#simulate temp data
temp_vector <- rep(seq(from = 0, to = 40), 4)

#keep salinity constant. often ppl will use the mean or median of a continuous value
sal_vector <- rep(mean(sharkriver2$Salinity,
                       na.rm = TRUE), 164)

#simulate seasonal data
season_vector <- c(rep('Winter', 41),
                   rep("Spring", 41),
                   rep('Summer', 41),
                   rep('Fall', 41))

#join with continuous data
data_pred <- data.frame(temp_vector, season_vector, sal_vector)
colnames(data_pred) <- c('WaterTemp', 'season', 'Salinity')

#use original model to predict outcomes
prediction <- predict(hypox_mod,
                      newdata = data_pred,
                      type = 'response',
                      se.fit = TRUE)

#pull out predictions
data_fig <- data.frame(data_pred,
                       prediction$fit,
                       prediction$se.fit)

#rename columns
colnames(data_fig) <- c('WaterTemp', 'season', 'Salinity', 'probability', 'se')

#graph the probabilities of hypoxia
ggplot(data_fig, aes(x = WaterTemp,
                     y = probability)) +
  geom_line(aes(colour = season)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = season), alpha = 0.3) +
  labs(x = 'Temperature (degrees C)', y = 'Probability of Hypoxia',
       color = 'Season', fill = 'Season') +
  theme_bw()

```

# Poisson Regression

## Step 1 define research question
Do water depth and habitat type significantly predict daisy parrotfish counts?

## Step 2 examine data and possible correlations

```{r examine again}
#filter data for our species
 parrotfish <- coralfish %>%
  filter(Species == 'Chlorurus sordidus')

#histograms
ggplot(parrotfish, aes(x = Count)) +
geom_histogram(color = 'black')

ggplot(parrotfish, aes(x = Depth_m)) +
geom_histogram(color = 'black')

ggplot(parrotfish, aes(x = Habitat)) +
geom_bar(color = 'black')

#scatterplots
ggplot(parrotfish, aes(x = Depth_m, y = Count)) +
geom_point(color = 'black')

ggplot(parrotfish, aes(x = Habitat, y = Count)) +
geom_point(color = 'black')


```

## Step 3 fit regression model

```{r regressions}
#transform chr to factor
parrotfish$Habitat <- factor(parrotfish$Habitat,
                             levels = c('Forereef',
                                        'Pass',
                                        'Lagoon',
                                        'Fringing Reef'))

#fit a poisson regression model
fish_mod <- glm(Count ~ Depth_m + Habitat,
                data = parrotfish,
                family = 'poisson')


```

## Step 4 evaluate model diagnostics

```{r eval model}
#examine model output
summary(fish_mod)

#examine residuals
plot(fish_mod)

#remove outliers
parrotfish2 <- parrotfish[-c(69, 87),]

#refit poisson model
fish_mod2 <- glm(Count ~ Depth_m + Habitat,
                data = parrotfish2,
                family = 'poisson')

#reexamine model output
summary(fish_mod2)

plot(fish_mod2)

#remove more outliers
parrotfish3 <- parrotfish2[-c(82, 99, 100),]

#refit model
fish_mod3 <- glm(Count ~ Depth_m + Habitat,
                data = parrotfish3,
                family = 'poisson')

#re-examine model
summary(fish_mod3)

plot(fish_mod3)

```

## step 5 communicate results 

```{r report}

#format results as gt table
tbl_regression(fish_mod3)

#export model rsults as a dataframe
fish_mod_summary <- tidy(fish_mod3,
                         exponentiate = TRUE, #reports exponentiated values
                         conf.int = TRUE) # adds confidence intervals

```

