---
title: "week3script"
author: "Summer Heschong"
date: "2025-01-29"
output: pdf_document
---
#load packages
```{r setup}
library(here)
library(tidyverse)
library(moments)

```

#Discrete Distributions (Bernoulli Trial and Poisson Distribution)
###Bernoulli Trial
```{r Bernoulli Trial: exact number of successes}
#probability of three successes 
dbinom(x = 3, size = 16, prob = 0.48)

#probability of 0-16 successes
prob <- dbinom(x = 0:16, size = 16, prob = 0.48)

#create a list of numbers of successes
success <- 0:16

#bind two columns together to create a dataframe
binom_df <- cbind(prob, success)

#create a fugure
ggplot(binom_df, aes(x = success,
                     y = prob)) +
  geom_bar(stat = "identity") +
  labs(x = 'X', 
       y = 'Probability',
       title = 'The probability of getting 0-16 successes out of 16 trials.')

```

```{r Bernoulli Trial: three or fewer successes}
#Cumulative probability using dbinom
sum(dbinom(x = 0:3, size = 16, prob = 0.48))

#cumulative probability using generated data
#adding together rows 1 through 4
sum(prob[1:4])

```

```{r Bernoulli Trial: k or fewer succeses}
#Probability of three of fewer successes
pbinom(q = 3, size = 16, prob = 0.48)

```
###Poisson distribution models counts of outcomes rather than modeling successes and failures
```{r Poisson Distribution}
#probability of 10 occurrences 
dpois(x = 10, lambda = 3)

#Probability of 3 or fewer occurrences
ppois(q = 3, lambda = 3)

```

#Assessing Normality
```{r assesing data for normality}
#consider information on how this data was collected
#?beaver1()

#visually inspect data
#View(beaver1)

#create a histogram of beaver body temperature
ggplot(beaver1, aes(x = temp)) + 
  geom_histogram() +
  labs(x = 'Beaver Body Temperature (C)',
       y = 'Count')

#create a Q-Q plot of beaver body temperature
ggplot(beaver1, aes(sample = temp)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = 'Theoretical Normal Distribution',
       y = 'Raw Temperature Values')
```

```{r skewness and kurtosis}
#calculate skewness of beaver body temperature
skewness(beaver1$temp)

#calculate kurtosis of beaver body temperature 
kurtosis(beaver1$temp)
```
the absolute value of skewness for beaver body temperature is > 0 and < 0.5 therefore it is approximately symmetrical

kurtosis for beaver body temperature is >3 therefore it is leptokurtic

#Continuous Distributions
pnorm() calculates area under the probability density function of a normal distribution to the left of a specified value of x
```{r using pnorm}
#probability of value < 1
pnorm(q = 1, mean = 2, sd = 3)

```
```{r save a dataset created using dnorm}
#prob of value < 1
pnorm(q = 1, mean = 2, sd = 3)

#create a list of values with lots of values
#in between whole numbers (so our curve looks nice and smooth)
values <- seq(-10, 14, length = 1000)

#probability densities for values -10 to 14
prob_norm <- dnorm(x = values, mean = 2, sd = 3)

#bind two columns together to create a dataframe
norm_df <- cbind(prob_norm, values)

#create a figure
ggplot(norm_df, aes(x = values,
                    y = prob_norm)) +
  geom_line() +
  labs (x= "X",
        y = 'Probability Density',
        title = 'The PDF of X ~ N(2,3).')
```
qnorm() returns the value of x that is the upper bound of a particular probabiltiy

```{r qnorm}
#value of 'x' for a given probability
qnorm(p = 0.80, mean = 2, sd = 3)

```


