---
title: "Lab Assignment 9"
author: "Summer Heschong"
date: "2025-03-25"
output: pdf_document
---
# (1) Logistic Regression
```{r setup}
#load packages
library(here)
library(tidyverse)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)

#load data
lizards <- read.csv(here('Data/Raw/jrn_lizard.csv'))

#filter dataset for side-blotched lizards
lizards <- lizards %>%
  filter(spp == 'UTST') %>%
  na.omit
```

## Step 1 - Define research question
Do snout-to-vent length, sex, and vegetation zone at time of capture significantly predict if a lizard tail is recorded as whole?

## Step 2 - Examine data
### a. Display distributions or raw counts of data
```{r histograms}
#create histograms displaying data

#snout-to-vent length
ggplot(lizards, aes(x = SV_length)) +
geom_histogram(color = 'black')

#sex
ggplot(lizards, aes(x = sex)) +
geom_bar ()

#vegetation zone
ggplot(lizards, aes(x = zone)) +
geom_bar ()

#lizard tail status
ggplot(lizards, aes(x = tail)) +
geom_bar ()
```

### b. Display relationships between predictor and outcome variables
```{r bar}
#create bar plots 

#snout-to-vent length
ggplot(lizards, aes(x = SV_length, fill = tail)) +
geom_bar (position = 'dodge')

#sex
ggplot(lizards, aes(x = sex, fill = tail)) +
geom_bar (position = 'dodge')

#zone
ggplot(lizards, aes(x = zone, fill = tail)) +
geom_bar (position = 'dodge')

```

## Step 3 - Fit regression model
```{r gen linear regression}
# change characters to factors
lizards$sex <- factor(lizards$sex, levels = c('J', 'F', 'M'))

lizards$zone <- factor(lizards$zone, levels = c('G', 'C', 'M', 'T'))

lizards$tail <- factor(lizards$tail, levels = c('B', 'W'))

#fit regression model
tail_mod <- glm(tail ~ SV_length + sex + zone,
                data = lizards,
                family = 'binomial')

```

## Step 4 - Evaluate model diagnostics
```{r diag}
#examine model output
summary(tail_mod)

#examine modified residuals plot
simulateResiduals(tail_mod) %>% plot()

```

## Step 5 - Interpret model and communicate results
### a. Report results
The results of a logistic regression suggest that the log-odds of a lizard tail being whole are greater with shorter snout-to-vent length (B = -0.04, p < 0.001). There was no significant relationship between the log-odds of a lizard tail being whole and the sex of the lizard relative to juvenile lizards (male: B = -0.21, p < 0.52; female: B = -0.33, p < 0.27). There was a significant increase in the log-odds of a lizard tail being whole when the lizard is captured in creosotebush shrubland (B = 0.51, p < 0.01) relative to grama grassland. However, lizards caught in mesquite duneland and tarbush shrubland did not have significantly different log-odds of a lizard tail being whole relative to those caught in grama grassland (zoneM: B = 0.28, p < 0.14; zoneT: B = 0.26, p < 0.22). 

### b. Provide 2 predictive figures
```{r results}
# simulate snout-to-vent length data
SV_length_vector <- rep(seq(from = 0, to = 99), 12)

#simulate sex data
sex_vector <- c(rep('M', 400),
                rep('F', 400),
                rep('J', 400))

#simulate zone data
zone_vector <- c(rep('G', 300),
                 rep('C', 300),
                 rep('M', 300),
                 rep('T', 300))

#join data
pred_data <- data.frame(SV_length_vector, sex_vector, zone_vector)
colnames(pred_data) <- c('SV_length', 'sex', 'zone')

#Use original model to predict outcomes
prediction <- predict(tail_mod,
                      newdata = pred_data,
                      type = 'respons',
                      se.fit = TRUE)
#pull out predictions
prob_data  <- data.frame(pred_data,
                        prediction$fit,
                        prediction$se.fit)
#rename columns
colnames(prob_data) <- c('SV_length', 'sex', 'zone', 'probability', 'se')

#Graph probabilities of lizard tail being whole
ggplot(prob_data, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = sex)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = sex), alpha = 0.3) +
  labs(x = 'Snout-to-Vent Length (mm)',
  y = 'Probability of a Lizard Tail Being Whole',
  caption = 'Predictive figure showing change in probability of a lizard tail being whole \n as lizard length increases, with lines colored by sex.')

ggplot(prob_data, aes(x = SV_length, y = probability)) +
  geom_line(aes(color = zone)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = zone), alpha = 0.3) +
  labs(x = 'Snout-to-Vent Length (mm)',
  y = 'Probability of a Lizard Tail Being Whole',
   caption = 'Predictive figure showing change in probability of a lizard tail being whole \n as lizard length increases, with lines colored by vegetation zone.')
``` 

# (2) Poisson Regression
## Step 1 - Define research question
Do season and plant species percent cover significantly predict lizard counts?
```{r setup2}
lizard_counts <- read.csv(here('Data/Raw/jrn_lizard_npp.csv'))
```

## Step 2 - Examine data and possible correlations
```{r examine}
#create histograms
ggplot(lizard_counts, aes(x = lizard_count)) +
geom_histogram(color = 'black')

ggplot(lizard_counts, aes(x = season)) +
geom_bar()

ggplot(lizard_counts, aes(x = BOER)) +
geom_histogram(color = 'black')

ggplot(lizard_counts, aes(x = LATR)) +
geom_histogram(color = 'black')

ggplot(lizard_counts, aes(x = PRGL)) +
geom_histogram(color = 'black')

ggplot(lizard_counts, aes(x = SCBR)) +
geom_histogram(color = 'black')

#Create scatter plots
ggplot(lizard_counts, aes(x = season, y = lizard_count)) +
geom_point(color = 'black')

ggplot(lizard_counts, aes(x = BOER, y = lizard_count)) +
geom_point(color = 'black')

ggplot(lizard_counts, aes(x = LATR, y = lizard_count)) +
geom_point(color = 'black')

ggplot(lizard_counts, aes(x = PRGL, y = lizard_count)) +
geom_point(color = 'black')

ggplot(lizard_counts, aes(x = SCBR, y = lizard_count)) +
geom_point(color = 'black')
```

## Step 3 - Fit regression model
```{r Poisson}
#transform characters to factors
lizard_counts$season <- factor(lizard_counts$season,
                                 levels = c('W', 'S', 'F'))

#fit poisson regression model
lizard_mod <- glm(lizard_count ~ season + BOER + LATR + PRGL,
                  data = lizard_counts,
                  family = 'poisson')
```

## Step 4 - Evaluate model diagnostics
```{r eval}
#examine model output
summary(lizard_mod)

#examine model residuals
plot(lizard_mod)

```

## Step 5 - Interpret the model and communicate the results
The results of a poisson regression suggest that summer (B = -1.01, p < 0.001) and fall (B = -0.30, p <0.001) have significantly lower side-blotched lizard counts relative to winter. Also lizard counts were found to be significantly greater as percent cover of black grama grass (B = 0.08, p < 0.001) and honey mesquite (B = 0.37, p < 0.001) increase, and percent cover of creosote bush decreases (B = -0.59, p < 0.001).

# (3) GitHub customization

https://github.com/srheschong



