---
title: "Lab Assignment 6"
author: "Summer Heschong"
date: "2025-02-19"
output: pdf_document
---
#Setup
```{r, Setup}
#load packages
library(here)
library(tidyverse)
library(moments)

#load data
vertebrate_data <- read.csv(here('Data/Raw/and_vertebrates.csv'))

```
# (1) Coastal Giant Salamander
##a. Are salamander data normally distributed?
```{r, assess salamander normality}
#filter dataset
salamander_data <- vertebrate_data %>%
  filter(species == 'Coastal giant salamander' & 
           (unittype == 'C' | unittype == 'P' | unittype == 'SC'))

#evaluate normality with figure
fig1 <- ggplot(salamander_data, aes(x = length_2_mm, fill = unittype)) +
  geom_histogram(bins = 45, binwidth = 6, color = 'black') +
  facet_wrap(~unittype) +
  labs(title = 'Coastal Giant Salamander Length by Habitat Type', 
       x = 'snout-to-tail length (mm)')
fig1

#calculate skewness
skewness(salamander_data$length_2_mm, na.rm = TRUE)

#calculate kurtosis
kurtosis(salamander_data$length_2_mm, na.rm = TRUE)

#not normal -> apply a log-transformation
salamander_data$length_2_mm <- log10(salamander_data$length_2_mm)

#re-evaluate normality

fig2 <- ggplot(salamander_data, aes(x = length_2_mm, fill = unittype)) +
  geom_histogram(bins = 45, binwidth = 0.1, color = 'black') +
  facet_wrap(~unittype) +
  labs(title = 'Coastal Giant Salamander Length by Habitat Type', 
       x = 'snout-to-tail length (log10 transformed)')
fig2

skewness(salamander_data$length_2_mm, na.rm = TRUE)

kurtosis(salamander_data$length_2_mm, na.rm = TRUE)

```
Answer: Based on fig1 the data looks like it is not normally distributed with a longer right tail. The skewness and kurtosis values confirm this. I applied a log-transformation and re-evaluated the data. In fig2 the data looks pretty normally distributed. The new kurtosis value (3) confirms this, and though the skewness value is not equal to 0 it is close.  

##b. Conduct a Bartlett's test for equal variance
```{r, Bartlett Test}
#perform Bartlett test
salamander_var <- bartlett.test(salamander_data$length_2_mm, 
                                salamander_data$unittype)
salamander_var

```
Answer: These data do not have approximately equal variances because the p-value for the Bartlett test was less than 0.05. 

#(2) Cutthroat Trout
##a. Are trout data normally distributed? 
```{r, assess trout normality}
#filter dataset
trout_data <- vertebrate_data %>%
  filter(species == 'Cutthroat trout')

#evaluate normality with figure
fig3 <- ggplot(trout_data, aes(x = length_1_mm, fill = reach)) +
  geom_histogram(bins = 55, binwidth = 4, color = 'black') +
  facet_wrap(~reach) +
  labs(title = 'Cutthroat Trout Length by Reach', 
       x = 'snout-to-fork length (mm)')
fig3

#calculate skewness and kurtosis
skewness(trout_data$length_1_mm, na.rm = TRUE)
kurtosis(trout_data$length_1_mm, na.rm = TRUE)

#not normal -> apply log-transformation
trout_data$length_1_mm <- log10(trout_data$length_1_mm)

#re-evaluate normality

fig4 <- ggplot(trout_data, aes(x = length_1_mm, fill = reach)) +
  geom_histogram(bins = 55, binwidth = 0.09, color = 'black') +
  facet_wrap(~reach) +
  labs(title = 'Cutthroat Trout Length by Reach', 
       x = 'snout-to-fork length (log10 transformed')
fig4

skewness(trout_data$length_1_mm, na.rm = TRUE)
kurtosis(trout_data$length_1_mm, na.rm = TRUE)


```
Answer: These data are not normally distributed as you can see in fig3. This is backed up by the kurtosis and skewness values. I applied a log-transformation to the data and re-evaluated. The data still do not appear to be normally distributed after the log-transformation. The skewness is less than 0 and the kurtosis is less than 3.