---
title: "Lab 4 Assignment"
author: "Summer Heschong"
date: "2025-02-05"
output: pdf_document
---
#Setup
```{r setup}
#load packages
library(here)
library(tidyverse)
library(dplyr)
library(vcdExtra)

#load in data set
urchin_data <- read.csv(here('Data/Raw/urchins.csv'))
```

#1. Data Tidying and Visualization

##a. Trim and filter dataset and convert frequency table to record of individual observations
```{r Wrangle Data}
#trim dataframe
urchin_data <- urchin_data %>%
  select(YEAR, MONTH, SITE, TREATMENT, SIZE, COUNT, COMMON_NAME)
  
#filter dataframe
red_urchin_data <- urchin_data %>% 
  filter(SITE == 'MOHK' & 
           COMMON_NAME == 'Red Urchin' & 
           YEAR >=2015 & YEAR <= 2016)

#convert to record of individual observations
red_urchin_data <- expand.dft(red_urchin_data, freq = 'COUNT')

#filter and convert dataframe for purple urchins for later (question 2.b.)
purple_urchin_data <- urchin_data %>% 
  filter(SITE == 'MOHK' & 
           COMMON_NAME == 'Purple Urchin' & 
           YEAR >=2015 & YEAR <= 2016)

purple_urchin_data <- expand.dft(purple_urchin_data, freq = 'COUNT')

```
##b. Visualize size distributions
```{r Visualize Data}
#create histogram
fig1 <- ggplot(red_urchin_data, aes(x = SIZE, fill = TREATMENT)) +
  geom_histogram(bins = 39, binwidth = 1, color = 'black') +
  facet_wrap(~YEAR) +
  labs(title = 'Size Distributions of Red Urchins in 2015 and 2016,
       Across Both Treatments in Mohawk Reef',
       x = 'Size (cm)')
fig1
```
#2. Confidence Intervals

##a. Calculate the 95%, two-sided CI for red urchin sizes in 2015 across both treatment types
```{r CI red urchins}
red_urchin_test <-t.test(red_urchin_data$SIZE,
                         subset = red_urchin_data$YEAR == 2015)
red_urchin_test
```
Answer: The 95% two-sided CI for red urchin sizes in 2015 across both treatment types is [8.1, 8.3]

##b. Calculate the 95%, two-sided CI for purple urchin sizes in 2016 across both treatment types
```{r CI purple urchins}
purple_urchin_test <- t.test(purple_urchin_data$SIZE,
                             subset = purple_urchin_data$YEAR ==2016)
purple_urchin_test
```
Answer: The 95% CI for purple urchin sizes in 2016 across both treatment types is [5.4, 5.5]

#3. One-sample Hypothesis Tests

##a. Is the mean red urchin size 9cm? Perform a one-sample test using 2015 data and one using 2016 data to investigate.
```{r One-Sample t-test}
#Null Hypothesis: the mean red urchin size is 9cm
#Alternative hypothesis: the mean red urchin size is not 9cm

#one-sample t-test for 2015
urchin_ttest_15 <- t.test(red_urchin_data$SIZE,
                         subset = red_urchin_data$YEAR==2015,
                         mu = 9,
                         alternative = 'two.sided')
urchin_ttest_15

#one-sample t-test for 2016
urchin_ttest_16 <- t.test(red_urchin_data$SIZE,
                         subset = red_urchin_data$YEAR==2016,
                         mu = 9,
                         alternative = 'two.sided')
urchin_ttest_16
```
Answer: For both 2015 and 2016 the p value is less than 0.05 therefore we reject the null hypothesis and retain the alternative hypothesis. The mean red urchin size for 2015 and 2016 is not 9cm.

#4. Two-Sample Hypothesis Test

##a. In 2015 was the mean red urchin size significantly different between the two treatments? In 2016?
```{r 2 Sample t-test Treatment Comparison}
#Prepare data for test:

#1. Create a dataset for Control 2015
red_urchin_C15_data <- red_urchin_data %>%
  filter(TREATMENT == 'CONTROL'&
         YEAR == 2015)

#2. Create dataset for Annual Removal 2015
red_urchin_AR15_data <- red_urchin_data %>%
  filter(TREATMENT == 'ANNUAL REMOVAL'&
         YEAR == 2015)

#3. Test for equal variances in 2015
var.test(x = red_urchin_C15_data$SIZE,
         y = red_urchin_AR15_data$SIZE)
#Result: variances are not equal

#1.Create a dataset for Control 2016
red_urchin_C16_data <- red_urchin_data %>%
  filter(TREATMENT == 'CONTROL'&
         YEAR == 2016)

#2,Create dataset for Annual Removal 2016
red_urchin_AR16_data <- red_urchin_data %>%
  filter(TREATMENT == 'ANNUAL REMOVAL'&
         YEAR == 2016)

#3. Test for equal variances in 2016
var.test(x = red_urchin_C16_data$SIZE,
         y = red_urchin_AR16_data$SIZE)
#Result: variances are not equal

#Perform 2-sample t-test:

#Null Hypothesis: the mean red urchin size is not significantly different
#between the two treatments
#Alternative Hypothesis: the mean red urchin size is significantly different
#between the two treatments

#2-sample t-test for 2015
urchin_2ttest_15 <- t.test(red_urchin_C15_data$SIZE,
                           red_urchin_AR15_data$SIZE)

urchin_2ttest_15

#2-sample t-test for 2016
urchin_2ttest_16 <- t.test(red_urchin_C16_data$SIZE,
                           red_urchin_AR16_data$SIZE)

urchin_2ttest_16
```
Answer: The mean red urchin size at Mohawk Reef for 2015 is not significantly different for the Control group (mean +- standard deviation = 8.18 +-1.38, n = 180) and the Annual Removal group (mean +- standard deviation = 8.21 +- 0.92, n = 175), as determined by a two-sample, two-sided t-test(t = -0.28, df = 313.51, p = 0.77, alpha = 0.05)

The mean red urchin size at Mohawk Reef for 2016 is significantly different for the Control group (mean +- standard deviation = 8.55 +-1.06, n = 101) and the Annual Removal group (mean +- standard deviation = 8.01 +- 0.85, n = 100), as determined by a two-sample, two-sided t-test(t = 4.01, df = 190.99, p = 8.60 e-5, alpha = 0.05)

##b. Was the mean red urchin size in the control data significantly different between the two years?
```{r 2 Sample t-test Year Comparison}
#First test for equal variances
var.test(x = red_urchin_C15_data$SIZE,
         y = red_urchin_C16_data$SIZE)
#Result: variances are not equal

#2-sample t-test for control data:

#Null Hypothesis: The mean red urchin size in the control group is not significantly
#different between 2015 and 2016
#Alternative Hypothesis: The mean red urchin size in the control group
#is significantly different between 2015 and 2016

urchin_2ttest_C <- t.test(red_urchin_C15_data$SIZE,
                          red_urchin_C16_data$SIZE)
urchin_2ttest_C
```
Answer: The mean red urchin size at Mohawk Reef for the Control Group is significantly different for 2015 (mean +- standard deviation = 8.18 +-1.38, n = 180) and 2016 (mean +- standard deviation = 8.55 +-1.06, n = 101), as determined by a two-sample, two-sided t-test(t = -2.54, df = 252.7, p = 0.01, alpha = 0.05)