---
title: "Lab 2 Assignment"
author: "Summer Heschong"
date: "2025-01-24"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---
#Setup: load packages and data
```{r setup}
library(here)
library(tidyverse)
library(viridis)
library(dplyr)

crab_data <- read.csv(here('Data/Raw/pie_crab.csv'))
```

#Part 1: Measures of location and data spread
```{r, Measures of location and data spread}
#a. What is the mean fiddler crab carapace size?

mean(crab_data$size)
#The mean fiddler crab carapace size is 14.66

#b. What is the median fiddler crab carapace size?

median(crab_data$size)
#The median fiddler crab carapace size is 14.44

#c. What is the standard deviation in fiddler crab carapace size? 

sd(crab_data$size)
#The standard deviation in fiddler crab carapace size is 3.5
```

#Part 2: Creating Histograms
```{r, Creating Histograms}
#a. Create histogram of fiddler crab carapace size
fig1 <- ggplot(data = crab_data, aes(x = size)) +
  geom_histogram(bins = 261, binwidth = 0.064) +
  labs(caption = 'Histogram of fiddler crab carapace size across sites')
print(fig1)

#To find the right number of bins I used 2n^1/3 or 2(392)^1/3,
#and for bin width I used range/#of bins

#b. Create histograms of fiddler crab carapace size for each of the 13 sites
fig2 <- ggplot(data = crab_data, aes(x = size)) +
  geom_histogram(bins=20, binwidth = 0.25) +
  facet_wrap(~site) +
  labs(caption = 'Histograms of fiddler crab carapace size by site')
print(fig2)
```
c. How do the raw values of fiddler crab carapace size in your first figure compare to the mean and median values you calculated above? Based on your second figure, do any sites display strong diﬀerences in their data distributions?

Answer: I think the raw values fit pretty well with the mean and median values I calculated above. In the first figure the values from around 12 - 16 all look pretty similar and there's a spike around 14. In the second figure all of the sites have pretty different data distributions. For example, at BC the data is pretty spread out across sizes, and at PIE the data is all to the left of size 15. 

#Part 3: Data Visualization - the Golden Rules
```{r, Data Visualization}
#a. Make a plot of your choice that follows the Golden Rules
fig3 <- ggplot(data = crab_data, aes(x = site, y = size)) +
  geom_point() +
  labs(caption = "Scatterplot of fiddler crab carapace size by site")
print(fig3)

#b. Create worst version of fig3
fig4 <- ggplot(
  data = crab_data, 
  aes(x = site, y = size, 
      color = size, size = 1 )) +
  geom_point() +
  facet_wrap(~site) +
  labs(caption = 'Carapace and island', x = 'x', y = 'y')
print(fig4)
```


